---
title: "Packages, Package management, and the Tidyverse"
subtitle: R for Psychology Research
author: "Marcus Lindskog, Docent"
date: "2019-09-08"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["rutgers", "marcus_uu.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "../templates/macros.js"
   #https://www.datamentor.io/r-programming/environment-scope/
---
```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(dev = 'svg')
options(device = function(file, width, height) {
  svg(tempfile(), width = width, height = height)
})
library(tidyverse)
```
# Overview
1. Packages

1. Package management

1. The `Tidyverse`
---
class: inverted, middle, center
# Packages
---

# What is a `package`?
1. A collection of **functions**, **data**, and **documentation**.
--

1. Extends the capabilities of base `R`.
--

1. Written by contributers to solve one or more statistical problem.
--

1. Available from CRAN or from the developer (e.g., GitHub).
---
# How do I find a useful package?

1. Google what it is you want to do!
--

1. Search https://Rdocumentation.org.

--

1. Search MRAN (the Microsoft R Application Network) https://mran.microsoft.com/packages

--

1. See the Trending R repositories list on GitHub.

--

1. See CRAN Tak views https://cran.r-project.org/web/views/

---
# How do I know which to use?

* Many packages do the same thing.

--

* How do I pick one that best does the job?

--

1. Try it out.

--

1. Ask friends.
--

1. Ask on social media, like twitter. The R-community is **very** helpful.

---
class: inverted, middle, center
# Package management
---
# Install a package.

* To install a package use:

--

```{r, eval = FALSE}
install.packages("packagename")

#For example

install.packages("nlme")
```

--

* To install the `nlm` packages that does mixed linear models.

--

* To check which packages that are currently installed.

--

```{r, eval = FALSE}
installed.packages()
```

--


---
# Load a packages

* Before you can use any of the functions or data in a package you need to make it available to your R session.

--

* This is done by loading your library.

--

* You can load your library from a script using:

--

```{r, eval = FALSE}
library(nlme)
```

--

* This is the recomended way, but R studio also allows you to load packages from the `Packages` tab.

---
# Load packages

* It is good practice to begin your analysis scripts by loading all packages that are needed for your analysis.

--

```{r load_packages, eval = FALSE}
#Load packages
library(plyr)
library(tidyverse)
library(knitr)
library(lubridate)
library(lme4)
```
---
# Detach

* If you, for some reason, want to remove a loaded library from your session, you need to detach it.

--

```{r, eval =FALSE}
library(BayesFactor)
detach("package:BayesFactor", unload = TRUE)
```
---
# Documentation

* If you want to know what a packages does, read the documentation.

--

* This can be found on https://Rdocumentation.org 

--

* And on https://cran.r-project.org

--

* The documentation often includes **Vignettes** and **Examples**.

---
# Updating

* Updating packages can sometimes be a hazzel.

--

* You should know that updating can change functionality.

--

* When you update your R version, all packges needs to be reinstalled. On Windows, this can be done using the `installr` package.

* On Mac, take some time to google a workflow.

--

* If you only want to check that all your packages are up to date, use:

```{r, eval = FALSE}
update.packages()
```

---
class: inverted, middle, center
# Tidyverse
---
# Tidyverse

* Preparing, wrangling, and visualizing data is a large part of doing statistics.

--

* There are many functions and packages in R to help you with that.

--

* However, the `Tidyverse` provides a collection of packages that work very well toghether and that are developed to solve these specific tasks.
---
# Tidy data

* The `Tidyverse` is developed by Hadly Wickham and builds on his idea of of **tidy data**.

--

* Tidy data
  1. Each variable forms a column.
  2. Each observation forms a row.
  3. Each type of observational unit forms a table.

--
* This is acctually very close to how we often think of a data set in psychology.

--

* The `Tidyverse` therefor works very well for our types of data.

---
# `Tidyverse` overview

* We will work with many of the packages from the `Tidyverse` later in the course.
--

* Here I will only give you a brief overview and some examples.

---
# `Tidyverse` packages

* The `Tidyverse` includes the following packages.

--

* Data Wrangling and Transformation
    * dplyr
    * tidyr 
    * stringr
    * forcats
--

* Data Import and Management
    * tibble
    * readr 

--

* Functional Programming
    * purrr
--

* Data Visualization and Exploration
    * ggplot2

---
# dplyr

* A very simple and agile package for data manipulation.
--

* Can use the pipe operator `%>%` to combine functions. This is very useful, as we will see later.

--
* Examples of functions from `dplyr`: 
  * `select()`: Select columns from your dataset
  * `filter()`: Filter out certain rows that meet your criteria(s)
  * `group_by()`: Group different observations together. 
  * `summarise()`: Summarise any of the above functions
  * `arrange()`: Arrange your column data
  * `join()`: Perform left, right, full, and inner joins in R
  * `mutate()`: Create new columns by preserving the existing variables

---
# tidyr

* Complements the `dplyr`packages by providing functinos to arrange columns.
--

* Examples of functions from `tidyr`
  * `gather()`: Gathers multiple columns and converts them into key-value pairs.
  * `spread()`: Takes two columns and spreads them into multiple columns.
  * `separate()`: Helps in separating or splitting a single column into numerous columns
  * `unite()`: Works opposite to the separate() function. Combines two or more columns into one


---
# stringr

* Working with strings often gives you a headache.

-- 
* stringer makes it a lot easier.

-- 
* Examples of functions from `stringr`

  * `str_sub()`: Extract substrings from a character vector
  * `str_trim()`: Trim white spaces
  * `str_remove()`: Removes a pattern from a character vector.
  * `str_to_lower/str_to_upper`: Converts into lower case or upper case
  
---
# Example

```{r, eval = FALSE}
od_data <- read_csv('analyses/data/OD_results.txt') %>%
  filter(!event_name == "-") %>%
  select(name, trialNr, event_name, AOI_name, AOI_value) %>%
  rename(ID = name) %>% rename(trial = trialNr) %>%
  mutate_at("event_name", str_remove, pattern = "a") %>%
  mutate(trial = as.numeric(trial),
         event_name = as.factor(event_name),
         AOI_value = as.numeric(AOI_value)) %>%
  spread(AOI_name, AOI_value) %>%
  mutate_at("ID", str_remove, pattern = ".csv") %>%
  separate(ID, c("ID", "R1", "Session", "Date",
                 "Time","X1","X2", "X3"), sep = "_") %>%
  select(-c("R1", "X1","X2", "X3")) %>%
  mutate_at("ID", str_remove, pattern = "S") %>%
  mutate(ID = as.numeric(ID),
         Session = as.factor(Session),
         Date = ymd(Date),
          PS = ifelse(Left_AOI < .200 | Right_AOI < .200  |
                        screen < 12.5*.25, NA, PS))
```

---
# readr

* Getting your data into R is sometimes a tricky task.

--
* The `readr` package provides functions that are very efficient to read in (and write) flat files.

--
* Examples of functions in `readr`

    * `read_delim()`: reads delimited files.
    * `read_csv()`: reads csv-files.
    * `write_csv()`: writes csv-files.

--
* You might also need to get data into R that come proprietary file formats (e.g., SAS, SPSS, Excel).

-- 
* To help you with this task we use the packages `haven` and `readxl`.


---
# tibble

* The `tibble` packages introduces a new data structure into R, the **tibble**.

--

* A tibble is a modern take on the data frame.

--

* It makes it easier and more consistent to work with tidy data in R.

---
# ggplot2

* R is perhaps best known for the graphs it can produce.

--

* `ggplot2` is one of the moste versitaile and powerful packages for graphing in R.

--

* It builds on Edward Tufte's **grammar of graphics**.

--

* Using a few basic elements, it can build nice looking graphs.

---
# Example
```{r}
library(tidyverse)

play_data <- cars
median_speed <- median(play_data$speed)
median_dist <- median(play_data$dist)
play_data$fast_long <- factor(ifelse(play_data$speed > median_speed & play_data$dist>median_dist, 1, 0),
                              labels = c("Shit", "Super"))
```
---
# Example

```{r}

my_plot <- ggplot(play_data, aes(x = speed, y = dist,
                                 color =  fast_long)) +
  geom_point()+
  labs(title = "Cars data",
       subtitle = "An investigation",
       x = "Speed",
       y = "Distance",
       color = "Type of car") + 
  theme_minimal() + 
  scale_color_manual(values = c("grey", "red"))+
  theme(panel.grid = element_blank(),
        axis.line = element_line(color = "grey"),
        legend.position = c(.10,.85)) + 
  geom_smooth(method="lm", color = "grey", 
              linetype = "dashed", se = FALSE) + 
  geom_hline(yintercept = median_dist,
             color = "grey", linetype = "dashed") + 
  geom_vline(xintercept = median_speed,
             color = "grey", linetype = "dashed") + 
  annotate("text", label = "Median Distance",
           x = 5, y = median_dist + 4, size = 4,
           colour = "grey")+
  annotate("text", label = "Median Speed", 
           x = median_speed + 1.5, y = 120,
           size = 4, colour = "grey")
```

---
```{r,echo=FALSE}
my_plot
```

---
class: inverted, middle, center
# That's all folks!

