--- 
title: "Reproducible Research Reports i R" 
subtitle: R for Psychology Research 
author: "Marcus Lindskog, Docent"
date: "2019-11-03"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["rutgers", "marcus_uu.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "../templates/macros.js"
--- 
```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(fig.dim=c(6, 6))
library(tidyverse)
library(knitr)
library(DT)
library(kableExtra)
```

# Overview
1. Not today!!

---
class: inverted, middle, center 
# Problem
---
# Questions

* How long would it take you to reproduce the results in your latest paper from raw data?
--

* If a reviewer asked you to add/drop a variable/condition in your analyses, how long would it take to get the new results section ready (including figures)?
--

* Could an independent researcher reproduce your results if she/he had your raw data and your paper?
---

# Reproducible Research

* If you feel a bit uncomfortable answering the previous questions, reproducible research reports might be for you.
--

* **Reproducible Research**: A workflow that integrates data and analyses in a way that makes it possible for an independent researcher to recreate the finding in a relatively straightforward way.
--

* **Reproducible Research Reports (RRRs) **: A report (e.g., your manuscript) that also integrates writing the report into the reproducible research workflow.
---

# The Case for RRRs

1. Adopting a workflow with RRRs increases transparency of your work. Others can easily see what you did.
--

1. Collaborators can easily double-check the results.
--

1. Producing RRRs makes it easy to revise and update analyses and visualizations based on feedback from, e.g., reviewers.
--

1. RRRs maintain a close coupling between data, analyses, and report. You will like that in 10 years time.
--

1. Using RRRs makes it easy to incorporate the same output from your analyses in a variety of formats (papers, presentations, posters, etc.)
---

# Why is preregistration and/or sharing data not enough?

* Preregistration and/or sharing data is GREAT!!
--

* But, there is a long way from raw data to final paper.
--

* With RRRs you can show exactly how you choose to walk down that path and make all of your decisions explicit.
---
class: inverted, middle, center
# Possible Solutions
---

# The Cumbersome way

* Write down in detail exactly all the steps you take from raw data to the finished Results section.
--

* Be prepared to share these notes with a reviewer/independent researcher.
--

* This is a good first step but takes a lot of time. Both to do at first and to use when reproducing.
---

# The Middle way
* Use software like **R**, **SPSS**, or **MATLAB** to create scripts that take you from raw data to results.
--

* This even better. Someone else can take your script and your data and reproduce your results, and all of your analysis decisions are available from the code.
--

* If you use **R** or **MATLAB** the scripts can also output all your figures, close to how they would appear in the paper.
--

* **Downside:** You still have to assemble your report, which might lead to "copy-paste-mistakes" and other issues.
---

# The Full Monty
1. Structure your projects in a consistent way on your computer.
--

1. Analyse your data by writing scripts in **R**.
--

1. Write your paper in *RStudio* using *R Markdown* and incorporate your analyses directly into your paper.
--

1. When your paper is finished, make your entire project publicly available (within the limits of your ethics approval) on, e.g., osf.io.
--

1. (Optional) Use version control to keep track of all changes you make, both to your analyses and to your paper.
---
class: inverted, middle, center
# But...
---

# What could be possible reasons for not trying out RRRs?

* You feel that Word is God's gift to man and can't consider ever writing in any other software.
--

* You enjoy spending hours trying to get **SPSS** to output within-subjects error bars on your split plot.
--

* You think that the best way to spend your Sunday is to figure out why you excluded participant 31 but not 38 form your analyses.
---
class: inverted, middle, center
# R Markdown
---
# What is R Markdown?

* A unified authoring framework that combines:
  * code
  * results
  * text.
--

* Combines various tools like, R, markdown, and pandoc to go from code to reports in wide range of formats (e.g., Word, PDF, etc.)
--

* R Markdown is available in the RStudio IDE.
---
class: inverted, middle, center
# Three types of content
---
# A *YAML* header

```{r, eval = FALSE}
---
title: "Untitled"
author: "Marcus Lindskog"
date: '2019-11-03'
output: pdf_document
---
```

--
```{r, eval = FALSE}
--- 
title: "Reproducible Research Reports i R" 
subtitle: R for Psychology Research 
author: "Marcus Lindskog, Docent"
date: "2019-11-03"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["rutgers", "marcus_uu.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "../templates/macros.js"
--- 
```

---
# Chunks of R code

- A code chunk starts with three backticks like ```` ```{r} ```` where `r` indicates the language name and ends with three backticks. You can write chunk options in the curly braces (e.g., set the figure height to 5 inches: ```` ```{r, fig.height=5} ````).



````markdown
`r ''````{r}
fit = lm(dist ~ speed, data = cars)
b   = coef(fit)
plot(cars)
abline(fit)
```
```
````

---

# Inline R code

- An inline R code expression starts with `` `r `` and ends with a backtick `` ` ``.

```{r}

x <- 5
```

````markdown
For a circle with the radius `r '\x60r x\x60'`,
its area is `r '\x60r pi * x^2\x60'`.
````
--

* For a circle with the radius `r x `,
its area is `r pi*x^2`.

--

````markdown
The formula for the area is $A = \pi \cdot r^2$. 
For a circle with the radius `r '\x60r x\x60'`,
its area is `r '\x60r pi * x^2\x60'`.
````
--
* The formula for the area is $A = \pi \cdot r^2$. 
For a circle with the radius `r x `,
its area is `r pi*x^2`.

---
# Chunk options
Option             | Run code | Show code | Output | Plots | Messages | Warnings 
-------------------|----------|-----------|--------|-------|----------|---------
`eval = FALSE`     | -        |           | -      | -     | -        | -
`include = FALSE`  |          | -         | -      | -     | -        | -
`echo = FALSE`     |          | -         |        |       |          |
`results = "hide"` |          |           | -      |       |          | 
`fig.show = "hide"`|          |           |        | -     |          |
`message = FALSE`  |          |           |        |       | -        |
`warning = FALSE`  |          |           |        |       |          | -

---
# New notebook and Compile

* Open a new *.Rmd* file from RStudio and choose your document type.

--

* To complie the notebook pres *Knit* or Cmd/Ctrl-Shift-K

---
# Formating 

* Text 

```

> *italic* or _italic_

> **bold** or __bold__

> `code` 

```

--
* Headers

```

> # 1st level

> ## 2nd level

> ### 3rd level 

```
---
# Tables

````markdown

Header A| Header B
--------|--------
Cont. A1| Cont. B1
Cont. A2| Cont. B2

````
--

Header A| Header B
--------|--------
Cont. A1| Cont. B1
Cont. A2| Cont. B2

--

* You can also produce nice looking tables using `kable` from the `knitr` package. Or functions from the `kableExtra` package.
---
# Bibliographies and Citations

* You might want to do some citations in your reports.
--

* Then you need your bibliography in a `.bib` file. Most citation managers will produce that for you.

--
```yaml
bibliography: rmarkdown.bib
csl: apa.csl
```
--

```markdown
Separate multiple citations with a `;`:
Blah blah [@smith04; @doe99].

You can add arbitrary comments inside the square brackets: 
Blah blah [see @doe99, pp. 33-35; also @smith04, ch. 1].

Remove the square brackets to create an in-text citation: 
@smith04 says blah, or @smith04 [p. 33] says blah.

Add a `-` before the citation to suppress the author's name: 
Smith says blah [-@smith04].
```
---
# An example

```{r}
x <- rnorm(100, .6, .2)
t_test_res <- t.test(x, mu = .5)
```
--

* A single sample t-test revealed that the mean *x* (*M* =  `r round(mean(x), 3)`) was significantly different (*t*( `r t_test_res$parameter` ) =  `r round(t_test_res$statistic,2)`, *p* = `r round(t_test_res$p.value,3)`) from .5.
--

````markdown
A single sample t-test revealed that the mean *x* 
(*M* =  `r '\x60r round(mean(x), 3)\x60'`) was 
significantly different (*t*( `r '\x60r t_test_res$parameter\x60'` ) =  
`r '\x60r round(t_test_res$statistic,2)\x60'`, *p* =
`r '\x60r round(t_test_res$p.value,3)\x60'`) from .5.
```` 

---
# Packages to help you...

* `papaja`: Reproducible APA manuscripts with R Markdown (https://crsh.github.io/papaja_man/)

--
* `kableExtra`: To get much more flexibility in formatting tables.

--

* `xaringan`: To make nice and easy presentations (https://github.com/yihui/xaringan/wiki). All presentations for this course was produced with xaringan.

--

* `citr`: So you get bibliographies into your markdown and can easily cite.

---
class: inverted, middle, center
# That's all folks!
---
