---
title: "Introduction to Data Visualization  with `ggplot2`"
subtitle: R for Psychology Research
author: "Marcus Lindskog, Docent"
date: "2019-09-23"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["rutgers", "marcus_uu.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "../templates/macros.js"
---
```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(fig.dim=c(6, 6))
library(tidyverse)
library(knitr)
library(DT)

```

# Overview
1. A note on Data Viz in R

1. A brief intro to Layered Grammar of Graphics

1. The data of the day

1. Creating a simple plot with `ggplot2`

1. Aesthetics

1. Facets

1. Geoms

1. Stats

1. Coordinate systems

1. Customizing plots

1. Week 5 - Exercises
---
class: inverted, middle, center
# A note on Data Viz in R
---

# A note on Data Viz in R

* A lot of the hype around R comes from the possibility to create nice looking graphs.
--

* There are many systems for making graphs in R.
--

* `ggplot2` is a very versatile system that will get you far.
--

* `ggplot2` also uses a coherent system that plays well with the rest of the `tidyverse`.
--

* A powerful system for Data Viz **cannot** make up for bad choices.
--

* A copy-paste approach will only get you so far.

---
class: inverted, middle, center
# Layered Grammar of Graphics
---

# `ggplot2` and the Grammar of Graphics

* `ggplot2` implements the *grammar of graphics (GoG)*

--

* GoG is a coherent system for describing and building graphs.

--

* In GoG, a plot is built out of one or more **layers**.
--

* Each layer consists of one or more components:
  * data and aesthetic mappings.
  * geometric objects
  * scales
  * facet specifications
  * statistical transformations
  * coordinate system.
--

* `ggplot2` makes it easy to build graphs with these components.

---
class: inverted, middle, center
# The data of the day
---
# The data of the day

* For the data viz examples that we will work with, we need a data set.
--

* I will use the `diamonds` data set that is included in the `ggplot2` package.
--

* An example of the data is shown in the table below.
--

```{r, echo = FALSE}
DT::datatable(
  head(diamonds[,1:7], 6),
  fillContainer = FALSE,options = list(pageLength = 10)
)
```

---
class: inverted, middle, center
# A simple plot with `ggplot2`
---
# Creating a ggplot
```{r, eval = FALSE}
# code chunk here
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price))
```

---
# Creating a ggplot
```{r}
# code chunk here
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price))
```
---
# What did the code do?

```{r, eval = FALSE}
# code chunk here
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price))
```
--

* `ggplot()` - Created a coordinate system that we can draw layers to. This actually creates an empty graph.
--

* The first argument to `ggplot(data = diamonds)` is the data that we want to draw.
--

* The second row adds a layer to the plot. In this case a layer of *points*.
--

* The `geom_point()` function takes a **mapping** argument that defines how variables in the data set are mapped to visual properties in the layer.

--

* In our example, the visual properties are the x- and y-coordinates.

---
class: inverted, middle, center
# Aesthetics
---
# Aesthetics mappings

* An aesthetic is a visual property of the objects in the plot.
--

* There are several aesthetics available in `ggplot2` and which are available also depends on the object you are trying to plot.
  * Size
  * Shape
  * Color
  * Alpha
  
--

* `ggplot2` will automatically assign a unique level of the aesthetic to each unique value of a variable (**scaling**)

--

* `ggplot2` will also add a legend to explain which level corresponds to which value.


---
# Example - color
```{r}
# code chunk here
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price, color = cut))
```
---
# Example - shape
```{r}
# code chunk here
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price, shape = cut))
```
---
# Example - alpha
```{r}
# code chunk here
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price, alpha = cut))
```
---
# Setting aesthetics explicitly

* It is also possible to set an aesthetic for an object explicitly. 
--

* This is then done **outside** of the `aes()` function.

--
```{r}
# code chunk here
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price), color = "blue")
```

---
class: inverted, middle, center
# Facets
---
# Facets

* Using aesthetics, we can add variables to our plot.
--

* Another way to add variables, particularly categorical ones, is to use **facets**.

--

* Facets is a way to split the data into subgroups, and plot each subgroup in a separate graph.

--

* Importantly, using facets keeps the scaling the same in all graphs, which helps when comparing data.

---

# Facet - `facet_wrap()`
```{r}
# code chunk here
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price)) + 
  facet_wrap(~cut, nrow = 3)
```
---
# Facet - `facet_grid()`
```{r}
# code chunk here
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price)) + 
  facet_wrap(color~cut)
```

---
# Facet - `facet_grid()`
```{r}
# code chunk here
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price)) + 
  facet_wrap(.~cut)
```

---
class: inverted, middle, center
# Geoms
---
# Geometric objects - Plot 1
```{r}
# code chunk here
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price))
```

---
# Geometric objects - Plot 2
```{r, warning=FALSE}
# code chunk here
ggplot(data = diamonds) +
  geom_smooth(mapping = aes(x = carat, y = price))
```
---
# Geometric objects

* Plot 1 and 2 describe the same data, but they are not identical.
--

* They use different **visual objects** to represent the data.
--

* In `ggplot2` visual objects are referred to as **geoms**.
--

* A geom is a geometrical object that the plot uses to represent the data.

--

* We change the geom(s) that we want to display by calling different geom functions.

--

* Every geom function takes a mapping argument. But not every aesthetic works with every geom.

---
# Geom functions - Examples

* `geom_histogram()`
* `geom_point()`
* `geom_boxplot()`
* `geom_smooth()` 
* `geom_bar()` 
* `geom_errorbar()`
* `geom_pointrange()`
* `geom_rect()`
* `geom_hline()`
* `geom_vline()`
* `geom_violin()`
* `geom_text()`

* See also https://www.ggplot2-exts.org for extension packages.

---

# Multiple geoms in the same plot
```{r}
# code chunk here
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price)) +
  geom_smooth(mapping = aes(x = carat, y = price))
```

---
# Geoms can share data and mapping
```{r}
# code chunk here
ggplot(data = diamonds, mapping = aes(x = carat, y = price)) +
  geom_point() +
  geom_smooth()
```

---
# And they can have their own
```{r}
# code chunk here
ggplot(data = diamonds, mapping = aes(x = carat, y = price)) +
  geom_point(aes(color = color)) +
  geom_smooth(aes(linetype = cut))
```

---
# Geoms do not have to plot the same data
```{r}
# code chunk here
ggplot(data = diamonds, mapping = aes(x = carat, y = price)) +
   geom_point(aes(color = color)) +
   geom_smooth(data = filter(diamonds, cut == "Fair"), se = FALSE)
```

---
class: inverted, middle, center
# Stats
---
# Statistical transformations
```{r}
# code chunk here
ggplot(data = diamonds) +
   geom_bar(aes(x = cut))
```

---
# Statistical transformations

* Some graphs, like scatter plots, plot the raw data from the data frame you pass to it.
--

* Other graphs, like bar graphs, calculate new values that are plotted. E.g.:
  * Bar charts, histograms, and frequency polygons bin the data and plot counts in each bin.
  * Smoothers, fit a model to the data and plot predictions from the model. 
  * Box plots compute robust summary statistics that are displayed.
--

* In `ggplot2` vocabulary, the algorithm that transforms raw data to the new values that are plotted is called a **stat**. 

--

* Every geom has a default stat. Conversely, every stat has a default geom.

--

* Hence, we can use geoms and stats interchangeably.

---
# Example 
```{r}
# code chunk here
ggplot(data = diamonds) +
   stat_count(aes(x = cut))
```

---
# Override the default stat
```{r}
# code chunk here
ggplot(data = diamonds) +
   geom_bar(mapping = aes(x = cut, y = ..prop.., group = 1))
```
---
class: inverted, middle, center
# Coordinate Systems
---
# Coordinate Systems

* `ggplot2` uses a Cartesian coordinate system by default.
--

* Most often, it is not necessary to change this default behavior. But sometimes it can be useful.
--

* We can also change the scale on which x, y are plotted.

---
# Coordinate system example

```{r}
# code chunk here
ggplot(data = diamonds) +
   geom_bar(aes(x = cut)) + 
  coord_flip()
```
---
# Coordinate system example - 2

```{r}
# code chunk here
ggplot(data = diamonds) +
   geom_bar(aes(x = cut)) + 
  coord_flip() + 
  coord_polar()
```
---
# Scale example

```{r}
# code chunk here
ggplot(data = diamonds) +
   geom_point(aes(x = carat, y = price)) + 
   scale_y_log10()
```
---
class: inverted, middle, center
# Customizing plots
---
# Positioning
```{r}
# code chunk here
ggplot(data = diamonds) +
   geom_bar(aes(x = cut, fill = cut)) 
```
---
# Positioning continued
```{r}
# code chunk here
ggplot(data = diamonds) +
   geom_bar(aes(x = cut, fill = clarity)) 
```
---
# Positioning continued further
```{r}
# code chunk here
ggplot(data = diamonds) +
   geom_bar(aes(x = cut, fill = clarity), position = "dodge") 
```
---
# Positioning continued further still
```{r}
# code chunk here
ggplot(data = diamonds) +
   geom_bar(aes(x = cut, fill = clarity), position = "fill") 
```
---
# Positioning - we need some more
```{r}
# code chunk here
ggplot(data = mpg) +
   geom_point(aes(x = displ, y = hwy)) 
```

---
# Positioning - we need just a little more
```{r}
# code chunk here
ggplot(data = mpg) +
   geom_point(aes(x = displ, y = hwy), position = "jitter") 
```
---
# Labels

```{r, eval = FALSE}
p <- ggplot(data = mpg) +
   geom_point(aes(x = displ, y = hwy), position = "jitter")

p + labs(x = "New x label", y = "New y label",
         title = "Add title above plot",
         subtitle = "Add a subtitle below title",
         caption = "Add a caption below plot") +
  annotate(geom = "text", x = 5, y = 40, label  = "Look at me up here")
```
---
# Labels

```{r, echo = FALSE}
p <- ggplot(data = mpg) +
   geom_point(aes(x = displ, y = hwy), position = "jitter")

p + labs(x = "New x label", y = "New y label",
         title = "Add title above plot",
         subtitle = "Add a subtitle below title",
         caption = "Add a caption below plot") +
  annotate(geom = "text", x = 5, y = 40, label  = "Look at me up here")
```
---
# Legends
```{r, eval = FALSE}
p <- ggplot(data = diamonds, mapping = aes(x = carat, y = price)) +
  geom_point(aes(color = color))

p + theme(legend.position = "bottom",
          axis.ticks.x = element_blank())
```
---
# Legends
```{r, echo = FALSE}
p <- ggplot(data = diamonds, mapping = aes(x = carat, y = price)) +
  geom_point(aes(color = color))

p + theme(legend.position = "bottom",
          axis.ticks.x = element_blank())
```
---
# Legends
```{r, eval = FALSE}
p <- ggplot(data = diamonds, mapping = aes(x = carat, y = price)) +
  geom_point(aes(color = color))

p + guides(color = "none")
```
---
# Legends
```{r, echo = FALSE}
p <- ggplot(data = diamonds, mapping = aes(x = carat, y = price)) +
  geom_point(aes(color = color))

p + guides(color = "none")
```
---
# Legends
```{r, eval = FALSE}
p <- ggplot(data = diamonds, mapping = aes(x = carat, y = price)) +
  geom_point(aes(color = color))

p + scale_color_discrete(name = "New Legend Title",
                        labels = c("AA", "BB", "CC",
                                   "DD", "EE", "FF",
                                   "GG"))
```
---
# Legends
```{r, echo = FALSE}
p <- ggplot(data = diamonds, mapping = aes(x = carat, y = price)) +
  geom_point(aes(color = color))

p + scale_color_discrete(name = "New Legend Title",
                        labels = c("AA", "BB", "CC",
                                   "DD", "EE", "FF",
                                   "GG"))
```
---
# Themes
* The basic appearance of a plot can be changed using **themes**.
--

* There are a number of built in themes in `ggplot2`.

--
* Other themes are provided in various packages.

--

```{r, eval = FALSE}
p <- ggplot(data = diamonds, mapping = aes(x = carat, y = price)) +
  geom_point(aes(color = color))

p + theme_bw()
```

---
# Themes 
```{r, echo = FALSE}
p <- ggplot(data = diamonds, mapping = 
              aes(x = carat, y = price)) +
  geom_point(aes(color = color))

p + theme_bw()
```

---
# Themes 
```{r, eval= FALSE}
p <- ggplot(data = diamonds, mapping = 
              aes(x = carat, y = price)) +
  geom_point(aes(color = color))

p + theme_bw() + 
   theme_bw()+
  theme(panel.grid = element_blank(),
        axis.ticks  = element_blank(),
        panel.border = element_blank())
```
---
# Themes 
```{r, echo= FALSE}
p <- ggplot(data = diamonds, mapping = 
              aes(x = carat, y = price)) +
  geom_point(aes(color = color))

p + theme_bw() + 
   theme_bw()+
  theme(panel.grid = element_blank(),
        axis.ticks  = element_blank(),
        panel.border = element_blank())
```
---
# Week 5 - Exercises

* Using the `diamonds` data set in `ggplot2` produce the following seven types of graphs using `ggplot2`.
--

* Scatter plot, Bar plot, Box plot, Density plot, Dot plot, Histogram, and Violin plot.
--

* For each graph, be creative with adding variables, changing positions, adding labels, adding legends and adding and customizing themes.
--

* As long as you produce the seven graphs in a meaningful way, there is no right or wrong. However, I want you to play with as many features of the plot as possible such that you get comfortable with manipulating various aspects of your graph.
--

* You get special extra points for being creative!!!! 
--

* If you produce boring graphs that don't explore the features of `ggplot2`, you will only be allowed to plot ugly graphs using SPSS for the rest of your life.

---
class: inverted, middle, center
# That's all folks!

