---
title: "R for Psychology Research"
subtitle: Week 8 - Exercises
author: "Marcus Lindskog"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# 1. Regression

* The file "Album_Sales.csv" contains a fictional data set (Field, 2017) with the following variables

  * **adverts -** Amount (in thousands of pounds) spent promoting the album before release.
  * **sales -** Sales (in thousands of copies) of each album in the week after release.
  * **airplay -** How many times songs from the album were played on a prominent national radio station in the week before release.
  * **attract -** How attractive people found the band's image (1 to 10).

  Field, A. P. (2017). *Discovering Statistics Using IBM SPSS Statistics* (5th ed.). London: Sage. [Fictional data set]

1. Calculate descriptiv statistics and zero order correlations between the four variables.

2. Create two different multiple linear regression models that both predict the **sales** variable, but with different predictors. Evaluate the models and model parameters.

3. Calculate standardized beta coefficients with 95% confidence intervals for the two models.

4. Compare the two models, which can better explain the data.

5. The data set `mtcars` is available in base R. Build a regression model that regresses the variables `cyl`, `disp`, and `hp` on `mpg`. What is the adjusted R-square of the model? aRe there any significant predictors?

6. Update (using `update()`) the model in 5 with a term that is the interaction between `drat` and `qsec`. What is the adjusted R-square of the model? aRe there any significant predictors? Does the new model provide a better fit to the data?

7. Calculate standardized beta coefficients with 95% confidence intervals for the model in 6.

8. Using the `mtcars` data set, build a regression model that regresses the variables `cyl`, `disp`, and `hp` on `am`. Note that `am` is a categorical variable. aRe there any significant predictors?


# 2. ANOVA

* The following code simulates data for a repeated measures design (you need to load `tidyverse`).
```{r, eval = FALSE}
Long <- rnorm(40, 123, 12)
Medium <- rnorm(40, 138, 12)
Short<- rnorm(40, 140, 22)
ID <- 1:40

within_subjects_data <- tibble(ID = factor(ID), Length_Long = Long,
                                Length_Medium = Medium,
                                Length_Short = Short)
```

9. Transform the data set into a format that you can analyse using `aov()`.

10. Run a repeated measures ANOVA to determine the effect of Length on the response variable.

* The following code simulates data for a 2x2 between subjects design (you need to load `tidyverse`).
```{r, eval = FALSE}
DV <- c(rnorm(40, 123, 12), rnorm(40, 138, 12))
IV1 <- rep(c("Long", "Short"), each = 40)
IV2 <- rep(rep(c("Wide", "Narrow"), each = 20), 2)
ID <- 1:80


two_by_two_design_data <- tibble(ID,DV, IV1, IV2)
```

11. Transform the data set into a format that you can analyse using `aov()`.

12. Run an ANOVA to determine the effect of IV1, IV2 and their interaction on DV.

# 3. Examination Exercises 
The solution to the exercises in this section should be handed in as a part of the examination. Your solution should be contained in a single R-script that is emailed to `marcus.lindskog@psyk.uu.se`. Your code should be well commented and easy to follow. Answers to any questions below should be written as a comment in the R-script after the code that produces the answer.

* The following code simulates data for a between subjects design (you need to load `tidyverse`). 
```{r, eval = FALSE}
Resp_A <- rnorm(40, 10, 1)
Resp_B <- rnorm(40, 11, 1.5)
Resp_C <- rnorm(40, 15, 3)

between_subjects_data <- tibble(Response_A = Resp_A,
                                Response_B = Resp_B, Response_C = Resp_C)
```
1. Determine the effect of Condition on Response and run appropriate Post Hoc tests.

2. Use `ggplot2` to visualize the results.

3. Transform the F-table into a tibble that could be used for further processing.

4. Assume that assumptions for parametric analyses aRe **not** met and run an appropriate test determine the effect of Condition on Response.


* The file "Alcohol_Attitudes.csv" contains a fictional data set (Field, 2017) with the following variables in which participants indicated their attitudes towards the drink on a scale from -100 to 100.
  * **beerpos** - Attitude towards beer after watching positive imagery.
  * **beerneg** - Attitude towards beer after watching negative imagery.
  * **beerneut**- Attitude towards beer after watching neutral imagery.
  * **winepos** - Attitude towards wine after watching positive imagery.
  * **wineneg** - Attitude towards a wine after watching negative imagery.
  * **wineneut** - Attitude towards a wine after watching neutral imagery.
  * **waterpos** - Attitude towards a water after watching positive imagery.
  * **waterneg** - Attitude towards a water after watching negative imagery.
  * **waterneut** - Attitude towards a water after watching neutral imagery.
  * **participant** - Participant ID.



  Field, A. P. (2017). *Discovering Statistics Using IBM SPSS Statistics* (5th ed.). London: Sage. [Fictional data set]
  
5. Run a 3x3 repeated measures ANOVA on to test the effect of **Drink** and **Imagery** on attitudes. For the ANOVA you should use functions available in Base R.

6. Plot the data using `ggplot2` and run appropriate Post Hoc Tests to disentangle any significant effects. For this, other packages than those in Base R may be used. 

* The file "Fat_Gain.csv" contains a data set that provides participants' non-deliberate exercise activities (NEA) and their fat gain after a period of 8 weeks of overeating. The variables aRe
  * **NEA -** Energy use (in calories) from non-deliberate exercise activities.
  * **Fat -** Fat gain (in kg) after 8 weeks of excessive calorie intake.

7. Use linear regression to determine if NEA reduces fat gain. Does NEA reduce fat gain?

8. Calculate standardized beta coefficients with 95% confidence intervals for the model in 7.



