--- 
title: "Common plots for Psychology using `ggplot2`" 
subtitle: R for Psychology Research 
author: "Marcus Lindskog, Docent"
date: "2019-10-09"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["rutgers", "marcus_uu.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "../templates/macros.js"
--- 
```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(fig.dim=c(6, 6))
library(tidyverse)
library(knitr)
library(DT)
```

# Overview 
1. Graph for t-tests.
1. Graph for correlations.
1. Graph for ANOVA-like designs.
1. Graph for multiple regressions.
---
class: inverted, middle, center 
# Graphs for t-tests 
---
class: inverted, middle, center 
# Single sample t-tests 
---
# We need some data 
```{r} 
# code chunk here 
t_test_data <- rnorm(100, .6, .3)
data_df <- tibble(dat = t_test_data) 
```
---
# Test the data against .5 
```{r, eval = FALSE}
ggplot(data = data_df, aes(x = 1, y = dat)) +
  geom_jitter(aes(x= 1.02), width = .01,
              alpha = .7) +
  stat_summary(fun.data = "mean_cl_boot",
               size = .5, colour = "red") +
  ylim(c(-.3,1.4))+
  xlim(c(0.8,1.2))+
  geom_hline(yintercept = .5, linetype =
"dashed") +
  theme_bw() +
  labs(y = "Proportion", x = "") +
  theme(axis.text.x = element_blank(),
        panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank())
``` 
---
#  Test the data against .5 
```{r, echo = FALSE}
ggplot(data = data_df, aes(x = 1, y = dat)) +
  geom_jitter(aes(x= 1.02), width = .01,
              alpha = .7) +
  stat_summary(fun.data = "mean_cl_boot",
               size = .5, colour = "red") +
  ylim(c(-.3,1.4))+
  xlim(c(0.8,1.2))+
  geom_hline(yintercept = .5, linetype =
"dashed") +
  theme_bw() +
  labs(y = "Proportion", x = "") +
  theme(axis.text.x = element_blank(),
        panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank())
``` 
---
class: inverted, middle, center 
# What did that code do?
---
class: inverted, middle, center 
# Two group t-tests 
---
#  We need some data 
```{r} 
group_1 <- rnorm(50, 100, 10) 
group_2 <- rnorm(50, 120, 20)
groups <- rep(c(1,2), each = 50) 
data_groups <- tibble(IQ = c(group_1, group_2),
group = groups) 
``` 
---
# Compare two groups 
```{r, eval = FALSE}
ggplot(data_groups,
       aes(x = factor(groups),
           y = IQ)) + 
  geom_jitter(width = .01, alpha = .7) +
  stat_summary(fun.data = "mean_cl_boot",
               size = .5, colour = "red")+
  ylim(c(50,175)) +
  theme_bw() +
  labs(y = "Proportion", x = "Group") +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank()) 
``` 
---
# Compare two groups 
```{r, echo = FALSE}
ggplot(data_groups,
       aes(x = factor(groups),
           y = IQ)) + 
  geom_jitter(width = .01, alpha = .7) +
  stat_summary(fun.data = "mean_cl_boot",
               size = .5, colour = "red")+
  ylim(c(50,175)) +
  theme_bw() +
  labs(y = "Proportion", x = "Group") +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank()) 
``` 

---
class: inverted, middle, center 
# What did that code do? 
---
class: inverted, middle, center
# Bi-variate correlations 
---
# We need some data 
```{r}
x <- rnorm(100, 100, 10) 
y <- rnorm(100, 100,14) 

corr_data <- tibble(x = x, y = y) 
```
---
# Bi-variate correlation 
```{r, eval = FALSE} 
ggplot(corr_data, aes(x = x, y = y)) +
  geom_point()+ 
  geom_smooth(method = "lm",
              se = FALSE)+
  ylim(c(55,135))+
  xlim(c(55,135))+
  theme_bw() +
  labs(y = "Y", x = "X") +
  theme(panel.grid = element_blank(),
        panel.border = element_blank())
```
---
# Bi-variate correlation 
```{r, echo = FALSE} 
ggplot(corr_data, aes(x = x, y = y)) +
  geom_point()+ 
  geom_smooth(method = "lm",
              se = FALSE)+
  ylim(c(55,135))+
  xlim(c(55,135))+
  theme_bw() +
  labs(y = "Y", x = "X") +
  theme(panel.grid = element_blank(),
        panel.border = element_blank())
```
---
class: inverted, middle, center
# What did that code do?
---
class: inverted, middle, center 
# ANOVA (like) designs 
---
# We need some data
```{r}
a_data <- rnorm(100, 100, 10) 
b_data <- rnorm(100, 110, 10) 
c_data <- rnorm(100, 120, 10) 
d_data <- rnorm(100, 130, 10) 
aov_data <- tibble(dat = 
                     c(a_data, b_data,
                       c_data, d_data),
                   group = factor(rep(c("A", "B",
                                        "C", "D") ,
                                      each = 100)))
```
---
# One factor ANOVA
```{r, eval = FALSE}

aov_data %>% 
  group_by(group) %>%
  summarise(mean = mean(dat, na.rm = TRUE),
            se = sd(dat)/sqrt(sum(!is.na(dat)))) %>%
  ggplot(aes(x = group, y = mean,
             color = group)) +
  geom_point() +
  geom_errorbar(aes(ymin = mean - 1.96*se,
                    ymax = mean + 1.96*se),
                width = .4) +
  ylim(c(90,140)) +
  theme_bw() +
  labs(y = "IQ", x = "Group") +
  theme(panel.grid = element_blank(),
        panel.border = element_blank()) +
  guides(color = "none") 
```
---
# One factor ANOVA
```{r, echo = FALSE}

aov_data %>% 
  group_by(group) %>%
  summarise(mean = mean(dat, na.rm = TRUE),
            se = sd(dat)/sqrt(sum(!is.na(dat)))) %>%
  ggplot(aes(x = group, y = mean,
             color = group)) +
  geom_point() +
  geom_errorbar(aes(ymin = mean - 1.96*se,
                    ymax = mean + 1.96*se),
                width = .4) +
  ylim(c(90,140)) +
  theme_bw() +
  labs(y = "IQ", x = "Group") +
  theme(panel.grid = element_blank(),
        panel.border = element_blank()) +
  guides(color = "none") 
```

---
class: inverted, middle, center 
# What did that code do? 
---
# We need some more data 

```{r} 
a_data <- rnorm(100, 100, 10)
b_data <- rnorm(100, 110, 10)
c_data <- rnorm(100, 120, 10)
d_data <- rnorm(100, 130, 10)
aov_data <- tibble(dat = c(a_data, b_data,
                           c_data, d_data),
                   group_1 = factor(rep(c("A", "B",
                                          "C", "D"),
                                        each = 100)),
                   group_2 = factor(rep(rep(c("Alpha", "Beta"),
                                            each = 50),4)))

```
---
# Two factor ANOVA (like) design 

```{r, eval = FALSE}
aov_data %>% 
  group_by(group_1, group_2) %>%
  summarise(mean = mean(dat, 
                        na.rm = TRUE),
            se = sd(dat)/sqrt(sum(!is.na(dat)))) %>%
  ggplot(aes(x = group_1, y = mean,
             color = group_2)) +
  geom_point(position = position_dodge(width = .3)) + 
  geom_errorbar(aes(ymin = mean - 1.96*se,
                    ymax = mean + 1.96*se), width = .4,
                position = position_dodge(width = .3)) +
  ylim(c(90,140))+
  theme_bw() + 
  labs(y = "IQ", x = "Group 1") + 
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        legend.position = "bottom") + 
  scale_color_discrete(name = "Group 2")
```
---
# Two factor ANOVA (like) design 
```{r, echo = FALSE}
aov_data %>% 
  group_by(group_1, group_2) %>%
  summarise(mean = mean(dat, 
                        na.rm = TRUE),
            se = sd(dat)/sqrt(sum(!is.na(dat)))) %>%
  ggplot(aes(x = group_1, y = mean,
             color = group_2)) +
  geom_point(position = position_dodge(width = .3)) + 
  geom_errorbar(aes(ymin = mean - 1.96*se,
                    ymax = mean + 1.96*se), width = .4,
                position = position_dodge(width = .3)) +
  ylim(c(90,140))+
  theme_bw() + 
  labs(y = "IQ", x = "Group 1") + 
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        legend.position = "bottom") + 
  scale_color_discrete(name = "Group 2")
```
---
class: inverted, middle, center 
# What did that code do?
---
class: inverted, middle, center
# Graph for multiple regression
---
# The code
```{r, eval = FALSE}
library(dotwhisker)

m1 <- lm(mpg ~ wt + cyl + disp + gear, data = mtcars)

dwplot(m1) + 
  theme_bw() + 
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        legend.position = "none") +
  geom_vline(xintercept = 0, color = "grey", linetype = "dashed")
  
```
---
# The graph
```{r, echo = FALSE, warnings = FALSE}
library(dotwhisker)

m1 <- lm(mpg ~ wt + cyl + disp + gear, data = mtcars)

dwplot(m1) + 
  theme_bw() + 
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        legend.position = "none") +
  geom_vline(xintercept = 0, color = "grey", linetype = "dashed")
  
```
---
# Examination

* The examination for this week is optional. 

1. Choose two graphs from published papers.
1. Recreate the graphs as closely as possible using `ggplot2`. The exercise is most fun if you chose graphs that are not so good.
1. Improve the graphs by removing elements and/or highlighting data to make them more clear.

* If you need to extract data from published graphs you can use the free web tool **WebPlotDigitizer**, which is available at: https://automeris.io/WebPlotDigitizer/

---
class: inverted, middle, center
# That's all folks!